Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2018-01-30 cabios-template2/bigsnpr-review-supp.tex
add andrey

@@

\newpage

@@

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{adjustbox}
\usepackage{multirow}

@@

\clearpage

@@

\clearpage
\subsection{Overview}

\begin{table}[!h]
\caption{Comparison of tools in low-level data management (first three rows) and high-level genetic analyses (the rest of rows). The bigstatsr and bigsnpr packages provide the most comprehensive and versatile infrastructure for data storage, processing and extensions; implement the main genetic analysis pipelines; and intrincically exploit memory-mapped files and parallel data processing.
Notes:
1: binary data format;
2: multi-core and cluster parallel computation;
3: support of imputed genotype data.
When comparing tools in terms of parallel computation, we do not consider the case of jobs manually distributed, e.g.\ splitting genotypes into batches. We are aware of the bitwise parallelism in PLINK 1.9, but this feature is used internally for specific computations; instead, we follow the official documentaion (https://www.cog-genomics.org/PLINK/1.9/parallel). The official documentation for the GenABEL project is at http://www.genabel.org/packages/ParallABEL.
GRM: Genetic Relationship Matrix; PCA: Principal Component Analysis; GWAS: Genome-Wide Association Study; PRS: Polygenic Risk Score.}
\begin{center}
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
\multirow{2}{*}{Toolkit/Feature} & bigstatsr~/ & \multirow{2}{*}{PLINK 1.9} & \multirow{2}{*}{GCTA} & SNPRelate~/ & \multirow{2}{*}{GenABEL} \\
 & bigsnpr & & & GWASTools &  \\
\hline
\hline
Proper genotype data format & FBM$^{2,3}$ & ped/bed$^2$ & - & GDS$^{2,3}$ & databel$^{2,3}$ \\ \hline
Data format extension, e.g. -omics & $\checkmark$ & - & - & - & $\checkmark$ \\ \hline
Vector/Matrix operations & $\checkmark^1$ & - & - & - & $\checkmark^1$ \\ \hline
Quality Control & $\checkmark$ & $\checkmark$ & - & $\checkmark$ & $\checkmark$ \\ \hline
GRM & $\checkmark^1$ & $\checkmark^1$ & $\checkmark^1$ & $\checkmark^1$ & $\checkmark^1$ \\ \hline
PCA & $\checkmark^1$ & $\checkmark$ & $\checkmark^1$ & $\checkmark^1$ & - \\ \hline
GWAS & $\checkmark^1$ & $\checkmark$ & $\checkmark^1$ & $\checkmark$ & $\checkmark^1$ \\ \hline
PRS & $\checkmark$ & $\checkmark$ & - & - & - \\ \hline
\end{tabular}
\end{adjustbox}
\end{center}
\label{tab:andrey}
\end{table}

### 2018-01-30 cabios-template2/bigsnpr-review.tex
add andrey

\title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France, \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

@@

We present the results of four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. PCA is a computationally intensive step of the GWAS, so that we further compare PCA methods on larger datasets.
Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only). 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

>>>>

\title[R packages for analyzing genome-wide data]{Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Sample \textit{et~al}.]{Florian Priv\'e\,$^{\text{\sfb 1,}*}$, Hugues Aschard\,$^{\text{\sfb 2,\sfb 3}}$, Andrey Ziyatdinov\,$^{\text{\sfb 3}}$ and Michael G.B. Blum\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France, \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.}

@@

We present the results of four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. PCA is a computationally intensive step of the GWAS, so that we further compare PCA methods on larger datasets.
Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only). 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer. We present a table summarizing the features of different software in table \ref{tab:andrey}.
